Categorical Foundations of E-Graphs With Binding

### 2025-07-10 LICS/sections/introduction.tex
fixed lines 450-452

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We propose to address this gap via a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over a category of semilattices (\textit{semilattice-enriched}).
The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
Enriched categories generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022,bonchi_string_2022-1,bonchi_string_2022-2,ghica_rewriting_2023,ghica_hierarchical_2023}.
The key ingredient of our categorical approach is the semilattice-enrichment.
By the latter we will mean an idempotent commutative semigroup given by a binary operator but will stick to the word semilattice as it is shorter.
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

Although our primary interest is theoretical, we believe that phrasing e-graphs in the more general language of category theory will open the door to new potential domains of applications that are already phrased in categorical language with a defined notion of rewriting and thus these applications will benefit from the mentioned optimisation technique.
% now use string diagrams and which require the application of optimisation techniques \cite{Selinger_2010, Hasegawa-traced}.
This may include areas as diverse as digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX}, functional programs~\cite{ghica-zanassi2023string}, or computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.
% However, in this work we offer little to the e-graph practitioner, as we focus exclusively on foundational aspects, introducing no new applications or improved e-graph algorithmics.

A final contribution of this work is giving a concrete combinatorial representation of string diagrams for semilattice-enriched categories in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs}.
We also give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022},  proving the representation \textit{sound and complete} with respect to our categorical semantics.  
As a corollary, this also provides a formalisation of the rewriting theory of e-graphs as encountered in the literature~\cite{EggPaper}.

@@

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}. 
We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  
Some further avenues for investigation, including potential applications, are given in the \nameref{sec:conclusion}.

% String diagrams
Our use of string diagrams is also central to the intuition behind our work \cite{Selinger_2010, joyal_geometry_1991}.  
There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica-zanassi2023string} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
% Hierarchical hypergraphs
Hierarchical (hyper-)graphs are also a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
% String diagram rewrites
Our work builds heavily on string diagram rewrite theory as developed in \cite{bonchi_string_2022,bonchi_string_2022-1, bonchi_string_2022-2}.

>>>>

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We propose to address this gap via a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over a category of semilattices (\textit{semilattice-enriched}).
The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
Enriched categories~\cite{Borceux_1994,Kelly2022BASICCO} generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3,fscd,ghica-zanassi2023string}.
The key ingredient of our categorical approach is the semilattice-enrichment.
By the latter we will mean an idempotent commutative semigroup given by a binary operator but will stick to the word semilattice as it is shorter.
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

Although our primary interest is theoretical, we believe that phrasing e-graphs in the more general language of category theory will open the door to new potential domains of applications that are induced by monoidal rather than algebraic theories with a defined notion of rewriting and thus these applications will benefit from the mentioned optimisation technique.
% now use string diagrams and which require the application of optimisation techniques \cite{Selinger_2010, Hasegawa-traced}.
This may include areas as diverse as digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX}, functional programs~\cite{ghica-zanassi2023string}, or computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.
% However, in this work we offer little to the e-graph practitioner, as we focus exclusively on foundational aspects, introducing no new applications or improved e-graph algorithmics.

A final contribution of this work is giving a concrete combinatorial representation of string diagrams for semilattice-enriched categories in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs} or monoidal e-graphs, as the former are a generalisation of e-graphs for monoidal theories.
We also give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3},  proving the representation \textit{sound and complete} with respect to our categorical semantics.  
As a corollary, this also provides a formalisation of the rewriting theory of e-graphs as encountered in the literature~\cite{EggPaper}.

@@

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, flatt2022small, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}. 
We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  
Some further avenues for investigation, including potential applications, are given in the \nameref{sec:conclusion}.

% String diagrams
Our use of string diagrams is also central to the intuition behind our work \cite{Selinger_2010, joyal_geometry_1991}.  
There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica-zanassi2023string} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
% Hierarchical hypergraphs
Hierarchical (hyper-)graphs are also a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
% String diagram rewrites
Our work builds heavily on string diagram rewrite theory as developed in \cite{bonchi_string_2022-1,bonchi_string_2022-2, bonchi_string_2022-3}.

### 2025-07-10 LICS_2025/monoidal_e_graphs/sections/introduction.tex
fixed lines 450-452

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}. 
We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  
We elaborate on some potential applications in the \nameref{sec:conclusion}.

>>>>

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, flatt2022small, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}. 
We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  
We elaborate on some potential applications in the \nameref{sec:conclusion}.

### 2025-07-10 POPL2026/paper/sections/categorical_semantics.tex
fixed lines 450-452

	\noindent
	plus the rules for equivalence of terms up to renaming of slots.
	Recalling that we use $+$ to encode this congruence of terms, the first three rules (plus reflexivity, symmetry, and transitivity) correspond to idempotence, commutativity and associativity of $+$ and the fact that $+$ respect the composition $(;)$.
	%  and tensoring $\otimes$ (traditional e-graphs do not have tensoring as they are defined for algebraic signatures).
	While the other rule is \textit{natural} to enforce when developing e-graphs with binding, we see that it is indeed the \textit{right} rule to enforce as we can see it follow by a purely categorical argument by the free enrichment as per the law \ref{law:distributivity}.
	We do not enforce any rules for renaming as string diagrams are essentially a nameless representation of terms.
	Moreover, some renaming rules are not consistent with string diagrams, as slotted e-graph try to represent, \textit{e.g.}, terms $(a + b)$ and $(c + d)$ as a single entity (up to renaming of free variables), which is not the case for string diagrams (unless we do closure conversion).

>>>>

	\noindent
	plus the rules for equivalence of terms up to renaming of slots.
	The earlier work of~\citet{ghica2024equivalencehypergraphsegraphsmonoidal} that introduced free semilattice enrichment to describe e-graphs using morphism terms used the $+$ of a semilattice to denote the equivalence of morphism subterms.
	For example, the e-graph (b) from \autoref{fig:e-graph-example} can be written as a term as 
	\[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\] where $+$ encodes the equivalence of $*$ and $<\!\!<$
	We can see that the first three rules (plus reflexivity, symmetry, and transitivity) above correspond to idempotence, commutativity and associativity of $+$ and the fact that $+$ respect the composition $(;)$.
	%  and tensoring $\otimes$ (traditional e-graphs do not have tensoring as they are defined for algebraic signatures).
	Then, while the binding congruence rule is \textit{natural} to enforce when developing e-graphs with binding, we see that it is indeed the \textit{right} rule to enforce as we can see it follow by a purely categorical argument by the free enrichment as per the law \ref{law:distributivity}.
	We do not enforce any rules for renaming as string diagrams are essentially a nameless representation of terms.
	Moreover, some renaming rules are not consistent with string diagrams, as slotted e-graph try to represent, \textit{e.g.}, terms $(a + b)$ and $(c + d)$ as a single entity (up to renaming of free variables), which in the case for string diagrams require an additional step of closure conversion.

### 2025-07-10 POPL2026/paper/sections/introduction.tex
fixed lines 450-452

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

\begin{figure*}[h!]
	\[
		\adjustbox{scale=0.6}{
			\tikzfig{figures/e-graph-example}
		}
	\]
	\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}}
	\label{fig:e-graph-example}
\end{figure*}

>>>>

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, flatt2022_small, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

\begin{figure*}[h!]
	\[
		\adjustbox{scale=0.6}{
			\tikzfig{figures/e-graph-example-2}
		}
	\]
	\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}}
	\label{fig:e-graph-example}
\end{figure*}
